Thesis Proposal

### 2019-10-07 ThesisProposal.Rmd
Redid the lit rev again to incorporate meng and rubin on assumptions (before research seminar feedback).

Word count: 711

@@

The numerous assumptions inherent to MI methods, can be roughly divided into assumptions about: 1) the *missingness mechanism*; 2)  the relation between observed and missing parts of the data (*imputation model*); and 3) the *non-response model*. By definition, assumptions cannot be verified on missing data. Therefore, proxy measures are used.

@@

The mechanism is assumed to be *missing at random*--i.e., the 'cause' of the missingness does not depend on the missing data itself [@rubin87]. Evaluation is done via sensitivity analyses to assess the robustness of inferences to violations. Some practical guidelines exist (e.g., [@nguy17]), but current MI software does not provide empirical researchers with this methodology.

The relation between observed and missing parts of the data can be evaluated by plotting conditional distributions [@abay08]. This procedure currently lacks an interactive interface and statistical tests to quantify the relations with covariates. Moreover, model fit could be assessed using *over-imputation* [@buur18] or *double robustness* [@bang05].

@@

The final assumption is met when the algorithm has converged to a stable distribution of MI values [@buur18]. Conventional measures to diagnose convergence--e.g., Gelman and Rubin's [-@gelm92] statistic $\widehat{R}$--do not suit multiply imputed data [@lace07]. Therefore, empirical researchers have to rely on a visual inspection procedure that is theoretically equivalent to $\widehat{R}$ [@whit11]. Not only is visually assessing convergence difficult for the untrained eye, it might also be futile. The convergence properties of MI algorithms lack scientific consensus [@taka17], and some default `MICE` techniques might not converge to stable distributions at all [@murr18]. Convergence diagnostics for MI methods have not been systematically studied [@buur18].

@@

A working beta version of `ShinyMICE` will be considered a sufficient milestone to proceed with writing a technical paper on the methodology and the software. I aim to submit this for publication in *Journal of Statistical Software*. Finally, ShinyMICE will be integrated into the existing MICE environment, and a vignette for applied researchers will be written.

>>>>

Word count: 739

@@

The validity of MI data depends on numerous assumptions that--by definition--cannot be verified from the incomplete data. Therefore, existing evaluation methods rely on proxy measures. For the following assumtions, no reliable proxy measures have been proposed and/or implemented in `MICE`: 1) *ignorability* of the *missingness mechanism*; 2) suitability of the *imputation models*; and 3) *compatibility* with the MI algorithm. 
<!-- Rubin (1987b, 160-66) subdivided the work needed to create imputations into three tasks. The modeling task chooses a specific model for the data, the estimation task formulates the posterior parameters distribution given the model and the imputation task takes a random draws for the missing data by drawing successively from parameter and data distributions @buur18 par 4.5.1 -->

<!-- account for the process that created the missing data, -->
<!-- preserve the relations in the data, and -->
<!-- preserve the uncertainty about these relations. -->

<!-- The modeling task chooses a specific model for the data -> assumes ignorability. -->

<!-- The estimation task computes the posterior distribution of the model parameters of the assumes correct specification of the model. ->

<!-- The imputation task draws plausible values from the conditional posterior distribution of the missing data given the estimated parameters -> assumes convergence in distribution -->

1. The missingness mechanism is said to be ignorable when the 'cause' of the missingness does not depend on the missing data itself [@rubin87]. Evaluation is usually done with sensitivity analyses to assess the robustness of inferences to violations. Some practical guidelines exist (e.g., [@nguy17]), but current MI software does not facilitate this methodology for empirical researchers.

@@

2. The imputation models should capture the relations between observed and missing parts of the data. These relations can be evaluated by plotting conditional distributions [@abay08]. Such visualizations are available in `MICE`, but subsequent statistical tests to quantify the relations with covariates are not yet provided. Additionally, there is potential to assess model fit by means of *over-imputation* [@buur18] or *double robustness* [@bang05].

@@

3. The third assumption is met when the MI algorithm converges to a stable distribution. Conventional measures to diagnose convergence--e.g., Gelman and Rubin's [-@gelm92] statistic $\widehat{R}$--do not suit multiply imputed data [@lace07]. Therefore, empirical researchers have to rely on a visual inspection procedure that is theoretically equivalent to $\widehat{R}$ [@whit11]. Not only is visually assessing convergence difficult for the untrained eye, it might also be futile. The convergence properties of MI algorithms lack scientific consensus [@taka17], and some default `MICE` techniques might not converge to stable distributions at all [@murr18]. Moreover, convergence diagnostics for MI methods have not been systematically studied [@buur18].

In short, the existing literature provides both possibilities and limitations to evaluating the validity of multiply imputed data. My aim is to (further) develop and implement the methodology.

@@

A working beta version of `ShinyMICE` will be considered a sufficient milestone to proceed with writing a technical paper on the methodology and the software. I aim to submit this for publication in *Journal of Statistical Software*. Finally, `ShinyMICE` will be integrated into the existing `MICE` environment, and a vignette for applied researchers will be written.
